\bf Estimating the number of clusters using cross-validation

### 2018-03-09 cvclust.tex
typos, update Rev 1 response

Gabriel's cross-validation the Owen and Perry generalization are designed
specifically for choosing the rank of a low-rank matrix approximation like the
singular value decomposition or the non-negative matrix factorization. While
for some purposes it makes sense to view $k$-means clustering as a sort of low
rank approximation to the data matrix, the discrete nature of clustering (each
observation gets assigned to one of $k$ possible clusters) imposes additional
structure on the problem that makes the prior procedures not directly
applicable.

@@

In many applications, including our application in the sequel to a yeast cell
cycle dataset, the number of observations $N$ is below the number of
features~$P$. When this happens, the sample covariance matrix $\hat \Sigma$
will not have full rank~$P$, and will be a poor estimate of the population
covariance matrix. In these situations, though, it will still be the case that
a decomposition $\hat \Sigma = \Gamma \Lambda \Gamma^\T$ exists with $\Gamma$
having orthonormal columns and $\Lambda$ being diagonal and positive-definite.
It will also be the case that $\Lambda^{-1/2}$ and hence $\tilde{\dataX}$
exist. What fails in these situation is that, since $\hat \Sigma$ is no longer
a good estimate of $\Sigma$, it will no longer be the case that our correction
reliably transforms the noise covariance to $I$. Still, in these situations,
applying the correlation correction may give better results than using the
original untransformed data. In practice we inspect the resulting estimate
$\hat k$ both with and without the correlation correction, and if the two
answers agree, we can be more confident in our answer.

>>>>

Gabriel's cross-validation and the Owen and Perry generalization are designed
specifically for choosing the rank of a low-rank matrix approximation like the
singular value decomposition or the non-negative matrix factorization. While
for some purposes it makes sense to view $k$-means clustering as a sort of low
rank approximation to the data matrix, the discrete nature of clustering (each
observation gets assigned to one of $k$ possible clusters) imposes additional
structure on the problem that makes the prior procedures not directly
applicable.

@@

In many applications, including our application in the sequel to a yeast cell
cycle dataset, the number of observations $N$ is below the number of
features~$P$. When this happens, the sample covariance matrix $\hat \Sigma$
will not have full rank~$P$, and will be a poor estimate of the population
covariance matrix. In these situations, though, it will still be the case that
a decomposition $\hat \Sigma = \Gamma \Lambda \Gamma^\T$ exists with $\Gamma$
having orthonormal columns and $\Lambda$ being diagonal and positive-definite.
It will also be the case that $\Lambda^{-1/2}$ and hence $\tilde{\dataX}$
exist. What fails in these situations is that, since $\hat \Sigma$ is no longer
a good estimate of $\Sigma$, it will no longer be the case that our correction
reliably transforms the noise covariance to $I$. Still, in these situations,
applying the correlation correction may give better results than using the
original untransformed data. In practice we inspect the resulting estimate
$\hat k$ both with and without the correlation correction, and if the two
answers agree, we can be more confident in our answer.
